---
title: "Introduction with a case study" 
subtitle: "Applying OmniScopeR to published AD multi-omics data (Knight ADRC, MSBB, ROSMAP)"
author: "Dany MUKESHA"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{OmniScopeR AD case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates a real-world use case of **OmniScopeR** using publicly available supplementary data from Eteleeb *et al.* (PLOS Biology, 2024) "Brain high-throughput multi-omics data reveal molecular heterogeneity in Alzheimer’s disease". The paper provides processed supplementary tables (XLSX) for differentially expressed genes, proteins, and metabolites (discovery cohort: Knight ADRC; replication cohorts: MSBB and ROSMAP). Those supplementary files are downloadable directly from the journal site. This vignette downloads those supplemental files, reads them, performs a small integration and visualization workflow using the `OmniScopeR` API (example functions), and produces simple interpretable outputs.

**Important:** some raw individual-level omics datasets referenced in the paper require controlled-access (NIAGADS / Synapse). Here we only use the public supplementary tables (S3, S8, S9, etc.) provided with the article which contain lists of differentially expressed features and summary tables.

# Preparation: packages and environment

```{r setup, message=FALSE, warning=FALSE}
# Lightweight packages used by the vignette. Install if missing.
needed <- c("readxl", "dplyr", "knitr")
newpkgs <- needed[!(needed %in% installed.packages()[,"Package"])]
if(length(newpkgs)) install.packages(newpkgs)

library(readxl)
library(dplyr)
library(knitr)
library(OmniScopeR)

# Load OmniScopeR (assumes it's installed or available in the library path)
if(!requireNamespace("OmniScopeR", quietly = TRUE)) {
  message("OmniScopeR not installed: vignette will call functions as if available.")
}
```

# Step 1: Download the public supplementary tables from the PLoS Biology article

We programmatically download the supplemental XLSX files linked in the article. These files are public and available via DOI redirects. We download 3 example files used in this vignette:

-   S3 Data: Differentially expressed genes in the Knight ADRC cohort (XLSX)
-   S8 Data: Differentially expressed proteins in the Knight ADRC cohort (XLSX)
-   S9 Data: Differentially abundant metabolites in the Knight ADRC cohort (XLSX)

```{r download-supp, message=FALSE, warning=FALSE}
# Create data directory
dir.create("inst/extdata/OmniScopeR_AD_case", recursive = TRUE, showWarnings = FALSE)
outdir <- "inst/extdata/OmniScopeR_AD_case"

# DOI URLs (as listed in the article supplementary index)
urls <- list(
  S3 = "https://doi.org/10.1371/journal.pbio.3002607.s026",
  S8 = "https://doi.org/10.1371/journal.pbio.3002607.s031",
  S9 = "https://doi.org/10.1371/journal.pbio.3002607.s032"
)

files <- list(
  S3 = file.path(outdir, "S3_DEGs_KnightADRC.xlsx"),
  S8 = file.path(outdir, "S8_Proteins_KnightADRC.xlsx"),
  S9 = file.path(outdir, "S9_Metabolites_KnightADRC.xlsx")
)

for(n in names(urls)){
  if(!file.exists(files[[n]])){
    message("Downloading ", n, " to ", files[[n]])
    tryCatch(
      download.file(urls[[n]], destfile = files[[n]], mode = "wb", quiet = TRUE),
      error = function(e) message("Download failed for ", urls[[n]], ": ", e$message)
    )
  } else message(files[[n]], " already exists: skipping download.")
}

list.files(outdir)
```

# Step 2: Read and inspect the tables

```{r read-supp, message=FALSE, warning=FALSE}
# Read sheets: these are often single-sheet xlsx files; adjust if necessary
deg_tab <- tryCatch(readxl::read_excel(files$S3, skip = 2), error = function(e) NULL)
prot_tab <- tryCatch(readxl::read_excel(files$S8, skip = 2), error = function(e) NULL)
met_tab  <- tryCatch(readxl::read_excel(files$S9, skip = 2), error = function(e) NULL)

# Quick checks
if(is.null(deg_tab)) stop("Failed to read S3 DEGs table: check download or file path.")

cat("S3 rows, cols:", dim(deg_tab), "\n")
cat("S8 rows, cols:", if(!is.null(prot_tab)) dim(prot_tab) else "NULL", "\n")
cat("S9 rows, cols:", if(!is.null(met_tab)) dim(met_tab) else "NULL", "\n")

# show first rows
knitr::kable(head(deg_tab, 6))
```

# Step 3: Prepare light-weight feature matrices for integration

These supplemental tables are lists of significant features with statistics (fold-change, p-value). They are not full count matrices. To demonstrate OmniScopeR's workflow we will construct a **toy sample-feature matrix** by treating each significant feature as a column and creating synthetic sample-level values using a small simulation reflecting the direction of change (up / down) reported in the table. This approach is only for demonstration of the visualization pipeline using real published feature lists: it does not attempt to reconstruct the original raw data.

```{r build-matrices, message=FALSE, warning=FALSE}
set.seed(42)
# Select top N features from each modality
N <- 50

deg_tab <- deg_tab |> dplyr::rename(pvalue = PValue)
prot_tab <- prot_tab |> dplyr::rename(pvalue = PValue)
met_tab <- met_tab |> dplyr::rename(pvalue = PValue)
top_genes <- deg_tab %>% arrange(pvalue) %>% head(N)
if(!is.null(prot_tab)) top_prot <- prot_tab %>% arrange(pvalue) %>% head(N) else top_prot <- NULL
if(!is.null(met_tab)) top_met  <- met_tab %>% arrange(pvalue) %>% head(N) else top_met <- NULL

# Build synthetic sample matrix: 100 samples (50 AD-like, 50 Control-like)
nsamp <- 100
pheno <- c(rep(1, 50), rep(0, 50)) # 1 = AD-like, 0 = control-like

# helper that simulates values based on logFC sign if available
sim_feature_matrix <- function(features_tbl, name_col = NULL, logfc_col = "logFC"){
  if(is.null(features_tbl)) return(NULL)
  if(is.null(name_col)) name_col <- colnames(features_tbl)[1]
  feat_names <- as.character(features_tbl[[name_col]])
  mat <- matrix(rnorm(nsamp * length(feat_names), sd = 1), nrow = nsamp, ncol = length(feat_names))
  # impose small mean shift for AD samples depending on logFC sign
  if(logfc_col %in% colnames(features_tbl)){
    shifts <- sign(as.numeric(features_tbl[[logfc_col]])) * 0.8
    for(i in seq_along(shifts)){
      mat[pheno == 1, i] <- mat[pheno == 1, i] + shifts[i]
    }
  }
  colnames(mat) <- feat_names
  as.data.frame(mat)
}

# Guess name columns in each table. If not present, use first column.
gene_name_col <- if("gene" %in% tolower(colnames(deg_tab))) colnames(deg_tab)[tolower(colnames(deg_tab))=="gene"][1] else colnames(deg_tab)[1]

genes_mat <- sim_feature_matrix(top_genes, name_col = gene_name_col, logfc_col = ifelse("logFC" %in% colnames(top_genes), "logFC", "log2FoldChange"))
prots_mat <- if(!is.null(top_prot)) sim_feature_matrix(top_prot, name_col = colnames(top_prot)[1], logfc_col = ifelse("logFC" %in% colnames(top_prot), "logFC", "log2FoldChange")) else NULL
mets_mat  <- if(!is.null(top_met))  sim_feature_matrix(top_met,  name_col = colnames(top_met)[1],  logfc_col = ifelse("logFC" %in% colnames(top_met), "logFC", "log2FoldChange")) else NULL

# Assign rownames as sample IDs
rownames(genes_mat) <- paste0("S", seq_len(nsamp))
if(!is.null(prots_mat)) rownames(prots_mat) <- rownames(genes_mat)
if(!is.null(mets_mat)) rownames(mets_mat) <- rownames(genes_mat)

# Inspect dimensions
list(genes = dim(genes_mat), proteins = if(!is.null(prots_mat)) dim(prots_mat) else NULL, metabolites = if(!is.null(mets_mat)) dim(mets_mat) else NULL)
```

# Step 4: Integrate with OmniScopeR and run the basic pipeline

> The following code calls the `OmniScopeR` functions defined in the package skeleton (omni_integrate, omni_causal, omni_visualize, omni_explain, omni_narrate). If you are running this vignette locally, make sure the installed OmniScopeR package implements those functions as per the package API.

```{r run-omniscope, message=FALSE, warning=FALSE}
if(!requireNamespace("OmniScopeR", quietly = TRUE)) {
  # If OmniScopeR is not installed, assume functions are available in the current environment
  omni_integrate <- get0("omni_integrate", ifnotfound = NULL)
  omni_causal     <- get0("omni_causal", ifnotfound = NULL)
  omni_visualize  <- get0("omni_visualize", ifnotfound = NULL)
  omni_explain    <- get0("omni_explain", ifnotfound = NULL)
  omni_narrate    <- get0("omni_narrate", ifnotfound = NULL)
}

# Basic checks
if(is.null(omni_integrate) || is.null(omni_causal)){
  message("OmniScopeR functions missing in the environment. Please install or source the package code to run the following pipeline.")
} else {
  # Integrate
  omni_obj <- omni_integrate(transcript = genes_mat, protein = prots_mat, metabolite = mets_mat)
  
  # Build simple causal (correlation-based) graph
  causal_graph <- omni_causal(omni_obj, threshold = 0.6)
  
  # Visualize (this calls base plotting from omni_visualize)
  omni_visualize(causal_graph, max_nodes = 60)
  
  # Explain: compute correlation with phenotype (synthetic cognitive score here)
  pheno_score <- rnorm(nsamp) + pheno*1.5 # synthetic continuous phenotype correlated with AD-like label
  importance_df <- omni_explain(omni_obj, phenotype = pheno_score)
  
  # Narrative output to console (or file)
  omni_narrate(omni_obj, causal_graph, importance_df, output_file = file.path(outdir, "OmniScopeR_AD_case_report.txt"))
}
```

# Step 5: Interpretation and next steps

This vignette used public supplementary tables (feature lists) from a PLOS Biology study and constructed a synthetic sample-level matrix so we could run an integration/visualization/explainability demo using `OmniScopeR`. This reproduces the *style* of analysis (integration, network identification, mapping of top features) using *real* published features while avoiding controlled-access raw data.

**Next steps a researcher can take using OmniScopeR with controlled-access data:**

-   Request access to the NIAGADS or Synapse datasets referenced in the article to obtain subject-level processed matrices; then re-run the same pipeline to produce biologically meaningful sample-level visualizations. (NIAGADS and Synapse both describe how to request access in the article Data Availability section.)
-   Replace the simple correlation-based causal graph with a more formal causal inference method (e.g., Bayesian networks) and add ML explainability overlays (SHAP, permutation importance) for models trained to predict cognitive decline.
-   Extend the visualization to an HTML Shiny dashboard for interactive exploration by clinicians.

# Session info

```{r session-info}
sessionInfo()
```
